Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-05-10 VegOzone.tex
Refine Section 4.1

We deduced the dominant species on-site at the NIBIO Environment Centre Svanhovd and from Fig.~\ref{fig:svanhovd_research_station} and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Default parameters for these species or PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as \emph{mapping manual} abbreviated as MM in the following. We base our parameterization of coniferous trees on the MM's Boreal Norway spruce and deciduous trees are derived from MM's silver birch. A comprehensive list of model parameters can be found in Supplement~Table~S1.
Initial test simulations with the MM PFTs showed unrealistically low $g_\mathrm{sto}$ in 2019 particularly for perennial grassland (Appendix~Fig.~\ref{fig:pody_mm_composit}). Only during an extended warm period in late July did we find substantial stomatal conductance. In ecological terms this would mean that there was almost no growth of grass in the summer of 2019 -- a prediction which is easily falsified by reality. We identified $f_\mathrm{T}$ as abnormally low, being the limiting factor of stomatal conductance. Therefore, we propose bespoke parameterizations which may describe an actual acclimation of these species to the climatic conditions in the target region.

Note, however, that these parameterizations represent hypothetical PFTs or species that have yet to be verified by experiments.

We hypothesize that perennial vegetation will likely prioritize their growth or fast adolescence and hence their carbon sequestration, during a short, subarctic growing season. Because the mapping manual version of the $\mathrm{DO_3SE}$ model currently does not simulate $A_\mathrm{n}$, we assume a first order proportionality between $A_\mathrm{n}$ and $g_\mathrm{sto}$ \citep{GCB:Medlyn2011} and effectively tune temperature and light response functions ($f_\mathrm{T}$ and $f_\mathrm{light}$) for higher $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$. We further assume that $f_\mathrm{VPD}$ and $f_\mathrm{SWP}$ suffice our biomes. The procedure is graphically illustrated in Fig.~\ref{fig:f_temp_grassland} for grassland. We aim to increase the enclosed area between probability density function (PDF) and the respective response function. We show two climatological reference periods 1990s and 2000s to indicate that the vegetation is subject to ongoing climate change and hence continuous acclimation. Resulting response functions for coniferous and deciduous trees are displayed in Appendix~Figs.~\ref{fig:f_temp_spruce}--\ref{fig:f_temp_birch}.

We define two different hypothetical temperature acclimation: \emph{subarctic} and \emph{cold}. The most extreme of the two acclimations is \emph{subarctic}. For either, we assume that the hypothetical species are more tolerant to cold conditions and shift $f_\mathrm{T}$ towards colder temperatures accordingly. We construct \emph{cold} as representative for a species that is more tolerant to cold temperatures, but slightly less efficient at warm temperatures compared to MM. In the same way, \emph{subarctic} refers to a hypothetical species which is very tolerant to cold but is sensitive to high temperatures, and most efficient at cool temperatures ($T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000). For perennial grassland the resulting response functions are depicted together with the PDF of temperature in Fig.~\ref{fig:f_temp_grassland}a).

Norway spruce was experimentally found to be active already at rather low air temperatures and could reach $60\,\unit{\%}$ photosynthetic activity as early as \unit{doy}~100 \citep{TB:Kolari2007, TP:Wallin2013}. Based on the time series of observed \chem{CO_2} uptake and temperature at two observation sites in Southern ($\rightarrow$ \emph{cold}) and Northern ($\rightarrow$ \emph{subarctic}) Finland \citep{TB:Kolari2007}, we estimate $10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$. We assume a similar acclimation for deciduous trees. All bespoke temperature parameters are listed in Table~\ref{tab:sensitivity_tests_temp}.\\

For an acclimation with respect to light conditions, we hypothesize that the light threshold for a $50\,\unit{\%}$ opening of plants' stomata may differ in subarctic species. We analytically derive the inverse function $f_{\mathrm{light},k}^{-1}$ for each PFT or species $k$
\begin{equation}
  \gamma_k(f_\mathrm{light}) := f_{\mathrm{light}, k}^{-1} = -\frac{\ln(1-f_\mathrm{light})}{\alpha_k}.
  \label{eq:inverse_function}
\end{equation}
First, we solve Eq.~(\ref{eq:inverse_function}) for the MM default value $\alpha_\mathrm{MM}^k$ for each PFT or species $k$ at $0.5$ 
\begin{equation}
  \gamma^k(0.5) = -\frac{\ln(0.5)}{\alpha_\mathrm{MM}^k},
  \label{eq:inverse_function_halfway}
\end{equation}
and define a variation $\gamma' = \gamma \cdot \eta$ with $\eta \in \{0.8, 1.2\}$. We invert Eq.~(\ref{eq:inverse_function}) for $\alpha$ and find $\alpha'(\gamma')$. The resulting functions for perennial grassland are shown in Fig.~\ref{fig:f_temp_grassland}b) (for deciduous and coniferous trees refer Appendix~Figs.~\ref{fig:f_temp_spruce}b)--\ref{fig:f_temp_birch}b)). All derived parameters are tabulated in Table~\ref{tab:sensitivity_tests_light}.

@@

As indicated above, a better acclimation will result in a higher average $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$. To quantify the resulting bespoke parameterizations, we propose the following metric. For a generic GS (May--August), we compute climatological average and standard deviation of $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ (Eq.~(\ref{eq:stomatal})) at noon (averaged over $11\,\unit{am}-1\,\unit{pm}$ local time) and in the morning (averaged over $5-9\,\unit{am}$ local time). We presume that relative stomatal conductance of around noon (highest light intensity) and in the early morning (lower intensities) are good proxies for \chem{CO_2} uptake efficiency. Our arbitrary optimization target is $60\,\unit{\%}$. Additional constrains are a small deviation between noon and morning and a low standard deviation indicating a higher robustness to interannual variability of growing conditions. For simplicity, we neglect the dependency on SWP in this regard. From the results shown in Fig.~\ref{fig:javis_func_opt_mean}, we identify \emph{subarctic}, PPFD0.8 as the best acclimation for all hypothetical species or PFTs alike. As expected due to the small adjustments, coniferous trees display the lowest deviation between the different types of parameterizations, while the divergence for perennial grassland is substantial as response to the stronger proposed temperature acclimation.

@@

Another important factor for ozone exposure metrics is the timing and length of the GS. Start ($A_\text{start}$) and end ($A_\text{end}$) of GS for each PFT are given in units of \unit{doy} (Table~\ref{tab:sensitivity_tests_gs}) and estimated as follows. Evergreen needle leaf trees show photosynthetic activity already at low temperatures \citep{TP:Wallin2013, TB:Kolari2007}. We use a MODIS (Aqua/Terra) retrieved net photosynthesis product for a $1\times 1\,\unit{km}$ patch centered at Svanhovd to determine $A_\text{start}$ and $A_\text{end}$. Net photosynthesis data can be approximated by a second order polynomial function (Fig.~\ref{fig:modis_Psn}). By calculating the roots of the fitted polynomial, we derive $A_\text{start} = 122\,\unit{doy}$ for 2018 and \unit{doy} 106 in 2019. $A_\text{end}$ amounts to \unit{doy} 261 and 274, respectively. This value will be used for all PFTs alike. The resulting growing season for coniferous trees in 2019 was one month longer then in 2018. For estimating $A_\text{start}$ of deciduous trees, we use the $5$ consecutive days above $5\,\unit{^\circ C}$ agricultural rule of thumb on gridded temperature data from SeNorge.no (Supplement~Fig.~\ref{fig:greening_season_change_Svanvik}). We find 129 and 130, respectively. For both years, these dates coincident with the first snow-free day at the meteorological weather station at Øvre Neiden (Sør-Varanger, NOR) which is relatively close-by.
For comparison, we would get $A_\mathrm{start} = 100\,\unit{doy}$ and $A_\mathrm{end} = 307\,\unit{doy}$ from the latitude model typically used for localizing $\mathrm{DO_3SE}$. By comparison, the latitude model indicates about a $2\,\unit{month}$ longer growing season. Due to a lack of quantitative field observation, we assume a latency period of about $1\,\unit{month}$ after snow-melt for perennial grassland.

@@

As pointed out by \citet{ACP:Bueker2012}, a very important factor in deriving a proper $\mathrm{POD_y}$ is the soil texture or how much water the soil can hold. As Svanhovd is located in the bed of the Pasvik river, we assume a sandy loam texture at $60\,\unit{cm}$ depth. The estimate of leaf boundary layer resistance is a function of wind speed which will be influenced by the tree height and leaf dimensions. Leaves of downy birch in northern Fennoscandia are smaller than southern specimens' with leaf dimensions of $(3.0\pm 0.5)\,\unit{cm}$. In addition, \citet[][p.~52]{NINA2004} indicate an average tree height of $13.5\,\unit{m}$ in the area. Both these factors will affect leaf boundary layer resistance, the smaller leaves would tend to reduce leaf boundary layer resistances whilst the shorter trees would reduce canopy roughness and increase canopy boundary layer resistances.

@@

We have modeled $\mathrm{POD_y}$ with a flux threshold $y=1\,\unit{nmol\,m^{-2}\,s^{-1}}$ per projected leaf area (PLA) for three natural/semi-natural vegetation types, deciduous trees, coniferous trees, and perennial grassland using both MM parameterizations and a variety of bespoke response function in $f_\mathrm{T}$ and $f_\mathrm{light}$, and localized GS (described in detail in Section~\ref{subsec:do3se_parameters}). The results are shown in Fig.~\ref{fig:pody_rel}. For comparison, critical levels for a $4\,\unit{\%}$ (deciduous), $2\,\unit{\%}$ (coniferous) \citep[refer to silver birch and Norway spruce in][]{ICP:MappingManual2017}, and $10\,\unit{\%}$ (perennial grassland, above ground) \citep{ESPR:Hayes2021} reduction in biomass are shown as horizontal dashed lines. Different years are color coded as previously. Simulations with MM GS, canopy, and leaf parameters are indicated with squared markers, bespoke GS with circles. The annotation $\pm$ refers to bespoke light response function threshold ($+\rightarrow\mathrm{PPFD1.2}$, $-\rightarrow\mathrm{PPFD0.8}$), while bespoke temperature response functions are indicated by categories on the x-axis. Closed/open symbols represent SWP sensitivity switched off/on in the model.

>>>>

We deduced the dominant species on-site at the NIBIO Environment Centre Svanhovd and from Fig.~\ref{fig:svanhovd_research_station} and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Default parameters for these species or PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as \emph{mapping manual} abbreviated as MM in the following. We base our parameterization of coniferous trees on the MM's Boreal Norway spruce and deciduous trees are derived from MM's silver birch. For a comprehensive list of model parameters consult Supplement~Table~S1.

Initial test simulations with the MM PFTs revealed an unrealistically low stomatal conductance at leaf-level ($G_\mathrm{sto}^\mathrm{leaf}$) in 2019 particularly for perennial grassland (Appendix~Fig.~\ref{fig:pody_mm_composit}). Substantial stomatal conductance occured only during an extended warm period in late July. In ecological terms, this means that there was almost no growth of grass in the summer of 2019 -- a prediction which is easily falsified by reality. We identified $f_\mathrm{T}$ as abnormally low, being the limiting factor of stomatal conductance in perennial grassland. Therefore, we saw the necessarity to propose bespoke parameterizations describing an acclimation of these species to the climatic conditions in the target region. Note, however, that these parameterizations represent hypothetical PFTs or species that have yet to be verified by experiments.

We hypothesize that perennial vegetation will likely prioritize their growth or fast adolescence and hence their carbon sequestration, during the short, subarctic growing season. Because the mapping manual version of the $\mathrm{DO_3SE}$ model currently does not simulate net photosynthesis ($A_\mathrm{n}$), we assume a first order proportionality between $A_\mathrm{n}$ and $g_\mathrm{sto}$ \citep{GCB:Medlyn2011} and effectively tune the temperature and light response functions ($f_\mathrm{T}$ and $f_\mathrm{light}$) for higher $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ (see Appendix~\ref{appendix:do3se_model} for details). We further assume that $f_\mathrm{VPD}$ and $f_\mathrm{SWP}$ suffice our biomes. The procedure is graphically illustrated in Fig.~\ref{fig:f_temp_grassland} for grassland. We aim to increase the enclosed area between probability density function (PDF) and the respective response function. We show two climatological reference periods 1990s and 2000s to indicate that the vegetation is subject to ongoing climate change and hence continuous acclimation. Resulting response functions for coniferous and deciduous trees are displayed in Appendix~Figs.~\ref{fig:f_temp_spruce}--\ref{fig:f_temp_birch}, respectively.

We define two different hypothetical temperature acclimation: \emph{subarctic} and \emph{cold}. For either, we assume that the hypothetical species are more tolerant to cold conditions and shift $f_\mathrm{T}$ towards colder temperatures accordingly. We construct \emph{cold} as representative for a species that is more tolerant to cold temperatures, but slightly less efficient at warm temperatures compared to MM. In the same way, \emph{subarctic} refers to a hypothetical species which is very tolerant to cold but is sensitive to high temperatures, and most efficient at cool temperatures ($T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000). The most extreme of the two acclimation types is \emph{subarctic}. For perennial grassland the resulting response functions are depicted together with the temperature PDF in Fig.~\ref{fig:f_temp_grassland}a).

Norway spruce was experimentally found to be active already at rather low air temperatures and can reach $60\,\unit{\%}$ photosynthetic activity as early as \unit{doy}~100 \citep{TB:Kolari2007, TP:Wallin2013}. Based on the time series of \chem{CO_2} uptake and temperature at observation sites in southern ($\rightarrow$ \emph{cold}) and northern ($\rightarrow$ \emph{subarctic}) Finland \citep{TB:Kolari2007}, we estimate the optimal temperature interval $10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$. We assume a similar temperature acclimation for deciduous trees. All bespoke temperature parameters are listed in Table~\ref{tab:sensitivity_tests_temp}.

With respect to light conditions, we hypothesize that the light threshold for a $50\,\unit{\%}$ opening of plants' stomata may differ in subarctic species compared to species in less extreme climates. Analytically, we derive the inverse function $f_{\mathrm{light},k}^{-1}$ of Eq.~(\ref{eq:flight}) for each PFT or species $k$
\begin{equation}
  \gamma_k(f_\mathrm{light}) := f_{\mathrm{light}, k}^{-1} = -\frac{\ln(1-f_\mathrm{light})}{\alpha_k}.
  \label{eq:inverse_function}
\end{equation}
First, we solve Eq.~(\ref{eq:inverse_function}) for the MM default value $\alpha_\mathrm{MM}^k$ for each PFT or species $k$ at $0.5$ 
\begin{equation}
  \gamma^k(0.5) = -\frac{\ln(0.5)}{\alpha_\mathrm{MM}^k},
  \label{eq:inverse_function_halfway}
\end{equation}
and define a variation $\gamma' = \gamma \cdot \eta$ with $\eta \in \{0.8, 1.2\}$. We invert Eq.~(\ref{eq:inverse_function}) for $\alpha$ and find $\alpha'(\gamma')$. The resulting functions for perennial grassland are shown in Fig.~\ref{fig:f_temp_grassland}b) (for deciduous and coniferous trees refer Appendix~Figs.~\ref{fig:f_temp_spruce}b)--\ref{fig:f_temp_birch}b)). All derived parameters are tabulated in Table~\ref{tab:sensitivity_tests_light}.

@@

As indicated above, a better acclimation to prevailing climate conditions will result in a higher average $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$. To quantify the resulting bespoke parameterizations, we propose the following metric. For a generic GS (May--August), we compute climatological average and standard deviation of $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ (Eq.~(\ref{eq:stomatal})) at noon (averaged over $11\,\unit{am}-1\,\unit{pm}$ local time) and in the morning (averaged over $5-9\,\unit{am}$ local time). We presume that relative stomatal conductance around noon (highest light intensity) and in the early morning (lower intensities) are good proxies for \chem{CO_2} uptake efficiency. Our arbitrary optimization target is $60\,\unit{\%}$. Additional constrains are a small deviation between noon and morning and a low standard deviation indicating a higher robustness to interannual variability of growing conditions. For simplicity, we neglect the dependency on SWP in this metric. From the results shown in Fig.~\ref{fig:javis_func_opt_mean}, we identify \emph{subarctic}, PPFD0.8 as the best acclimation for all hypothetical species or PFTs alike. As expected due to the small adjustments, coniferous trees display the lowest deviation between the different types of acclimation, while the divergence for perennial grassland is substantial as response to the stronger proposed temperature acclimation.

@@

Another important factor for ozone exposure metrics is the timing and length of the GS. Start ($A_\text{start}$) and end ($A_\text{end}$) of GS for each PFT are given in units of \unit{doy} (Table~\ref{tab:sensitivity_tests_gs}) and estimated as follows. Coniferous trees show photosynthetic activity already at low temperatures \citep{TP:Wallin2013, TB:Kolari2007}. We use a MODIS (Aqua/Terra) retrieved net photosynthesis product for a $1\times 1\,\unit{km}$ patch centered at Svanhovd to determine $A_\text{start}$ and $A_\text{end}$. The net photosynthesis data can be approximated by a second order polynomial function (Fig.~\ref{fig:modis_Psn}). By calculating the roots of the fitted polynomial, we derive $A_\text{start} = 122\,\unit{doy}$ for 2018 and \unit{doy} 106 in 2019. $A_\text{end}$ amounts to \unit{doy} 261 and 274, respectively. This value will be used for all PFTs alike. The resulting growing season for coniferous trees in 2019 was one month longer then in 2018. For estimating $A_\text{start}$ of deciduous trees, we use the $5$ consecutive days above $5\,\unit{^\circ C}$ agricultural rule of thumb on gridded temperature data from SeNorge.no (Supplement~Fig.~\ref{fig:greening_season_change_Svanvik}). We find 129 and 130, respectively. For both years, these dates coincide with the first snow-free day at the meteorological weather station at Øvre~Neiden (Sør-Varanger, NOR) which is relatively close-by.
For comparison, we would get $A_\mathrm{start} = 100\,\unit{doy}$ and $A_\mathrm{end} = 307\,\unit{doy}$ from the latitude model typically used for localizing $\mathrm{DO_3SE}$. The latitude model indicates a $2\,\unit{month}$ longer growing season. Due to a lack of quantitative field observation, we assume a latency period of $1\,\unit{month}$ after snow-melt for perennial grassland.

@@

As pointed out by \citet{ACP:Bueker2012}, a very important factor in deriving a proper $\mathrm{POD_y}$ is the soil texture or how much water the soil can hold. As Svanhovd is located in the bed of the Pasvik river, we assume a sandy loam texture at $60\,\unit{cm}$ depth. The estimate of leaf boundary layer resistance is a function of wind speed which will be influenced by the tree height and leaf dimensions. Leaves of downy birch in northern Fennoscandia are smaller than southern specimens' with leaf dimensions of $(3.0\pm 0.5)\,\unit{cm}$. In addition, \citet[][p.~52]{NINA2004} indicate an average tree height of $13.5\,\unit{m}$ in the area. Both factors will affect leaf boundary layer resistance, the smaller leaves would tend to reduce leaf boundary layer resistances whilst the shorter trees would reduce canopy roughness and increase canopy boundary layer resistances and hence the uptake of \chem{O_3}.

@@

We have modeled $\mathrm{POD_y}$ with a flux threshold $y=1\,\unit{nmol\,m^{-2}\,s^{-1}}$ per projected leaf area (PLA) for the three natural/semi-natural vegetation types, deciduous trees, coniferous trees, and perennial grassland using the generic MM parameterization and our bespoke temperaure (\emph{cold}, \emph{subarctic}) and light (PPFD0.8, PPFD1.2) acclimation types, and localized GS. The results are shown in Fig.~\ref{fig:pody_rel}. For comparison, critical levels for a $4\,\unit{\%}$ (deciduous), $2\,\unit{\%}$ (coniferous) \citep[refer to silver birch and Norway spruce in][]{ICP:MappingManual2017}, and $10\,\unit{\%}$ (perennial grassland, above ground) \citep{ESPR:Hayes2021} reduction in biomass are shown as horizontal dashed lines. 2018 is color-coded in pink and 2019 in purple. Simulations with MM GS, canopy, and leaf parameters are indicated with squared markers, bespoke GS with circles. The annotation $\pm$ refers to the bespoke light response function ($+\rightarrow\mathrm{PPFD1.2}$, $-\rightarrow\mathrm{PPFD0.8}$), while bespoke temperature response functions are indicated by categories on the x-axis. Closed/open symbols represent SWP sensitivity switched off/on in the model.
